---
title: "Data_preprocessing"
output: html_document
date: "2024-08-04"
---

Set up environment
```{r setup, include=FALSE}
source("src/environment_set.R") #Load functions and packages
set.seed(123) #For reproducibility
```

### LUNG PREDICT

Load data
```{r}
counts = read.csv("Counts_LungPredict.csv", row.names = 1) 
traitData_LP = read.csv("ClinicalData_LungPredict.csv", row.names = 1)
```

Keep only ENSEMBL IDs
```{r}
m2 <- do.call(rbind, strsplit(rownames(counts), split="_", fixed = TRUE))
counts <- as.matrix(counts)
rownames(counts) = m2[,1]
```

Extract only coding genes
```{r}
library(biomaRt)

mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

gene_ids <- rownames(counts)

# Get gene type annotations
annotations <- getBM(attributes = c("ensembl_gene_id", "gene_biotype"), 
                     filters = "ensembl_gene_id", 
                     values = gene_ids, 
                     mart = mart)

# Filter annotations to keep only coding genes
coding_genes <- annotations %>%
  filter(gene_biotype == "protein_coding") %>%
  pull(ensembl_gene_id)
 
# Subset the count matrix to keep only coding genes
counts <- counts[rownames(counts) %in% coding_genes, ]
```

Change to SYMBOL
```{r}
# Extract gene symbols 
entrz <- AnnotationDbi::select(org.Hs.eg.db, keys = rownames(counts), columns = "SYMBOL", keytype = "ENSEMBL") 

# Choose first gene symbol whenever one ENSEMBL ID have multiple Gene SYMBOLS
entrz <- entrz %>%
  group_by(ENSEMBL) %>%
  slice(1) %>%
  ungroup()

# Final count matrix
counts_LP = counts %>%
  data.frame() %>%
  mutate(Genes_symbol = entrz$SYMBOL) %>%
  na.omit(.) %>% #Remove genes with NO gene symbol
  remove_rownames() %>%
  distinct(Genes_symbol, .keep_all = TRUE) %>% #Keep only first gene symbol if duplicated
  column_to_rownames("Genes_symbol") 

```

### VANDERBILT

Load data
```{r}
counts = read.csv("Counts_Vanderbilt.csv", row.names = 1) 
traitData_Van = read.csv("ClinicalData_Vanderbilt.csv", row.names = 1)
```

Extract only coding genes
```{r}
library(stringr)
rownames(counts) <- str_replace(rownames(counts),
                        pattern = ".[0-9]+$",
                        replacement = "")

library(biomaRt)

mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

gene_ids <- rownames(counts)

# Get gene type annotations
annotations <- getBM(attributes = c("ensembl_gene_id", "gene_biotype"), 
                     filters = "ensembl_gene_id", 
                     values = gene_ids, 
                     mart = mart)

# Filter annotations to keep only coding genes
coding_genes <- annotations %>%
  filter(gene_biotype == "protein_coding") %>%
  pull(ensembl_gene_id)

counts = counts[rownames(counts) %in% coding_genes,]

# Change to gene SYMBOL
entrz <- AnnotationDbi::select(org.Hs.eg.db, keys = rownames(counts), columns = "SYMBOL", keytype = "ENSEMBL") %>%
  distinct(SYMBOL, .keep_all=T) %>%
  filter(ENSEMBL %in% rownames(counts)) %>%
  group_by(ENSEMBL) %>%
  slice(1) %>%
  ungroup()

counts = counts %>%
  rownames_to_column("Genes") %>%
  filter(Genes %in% entrz$ENSEMBL) %>%
  mutate(Symbol = entrz$SYMBOL) %>%
  na.omit() %>%
  remove_rownames() %>%
  column_to_rownames("Symbol") %>%
  dplyr::select(-Genes)

# Get gene type annotations
annotations_symbol <- getBM(attributes = c("hgnc_symbol", "gene_biotype"),
                     filters = "hgnc_symbol",
                     values = rownames(counts),
                     mart = mart)

# Filter annotations to keep only coding genes
coding_genes <- annotations_symbol %>%
  filter(gene_biotype == "protein_coding") %>%
  pull(hgnc_symbol)

counts_Van = counts[rownames(counts) %in% coding_genes,]
```

Keep same genes
```{r}
counts_LP = counts_LP[which(rownames(counts_LP)%in%rownames(counts_Van)),]
counts_Van = counts_Van[which(rownames(counts_Van)%in%rownames(counts_LP)),]
```

Normalize counts
```{r}
counts.norm.LP = ADImpute::NormalizeTPM(counts_LP, log=T) 
counts.norm.Van = ADImpute::NormalizeTPM(counts_Van, log=T) 
counts = cbind(counts.norm.LP, counts.norm.Van)
```

Join traitdata
```{r}
traitData_Van = traitData_Van %>%
  mutate("Batch" = "Van")

traitData_LP = traitData_LP %>%
  mutate("Batch" = "LP")

traitData = traitData_LP %>%
  bind_rows(traitData_Van) %>%
  dplyr::select(where(~!any(is.na(.))))
```

Only early stage
```{r}
traitData_early = traitData %>%
  filter(Stages_simplified %in% c("I", "II", "Stage_I", "Stage_II"))
```

PCA analysis with normalized counts
```{r}
compute_pca_analysis(counts[,rownames(traitData_early)], traitData_early, 'Batch')
```

Calculate TFs activity
```{r}
tfs_LP = compute.TFs.activity(counts.norm.LP)
tfs_Van = compute.TFs.activity(counts.norm.Van)
tfs = rbind(tfs_LP, tfs_Van)
```

PCA analysis with TFs
```{r}
compute_pca_analysis(t(tfs[rownames(traitData_early),]), traitData_early, 'Batch')
```

Performed deconvolution
```{r}
deconv_LP = compute.deconvolution(counts.norm.LP, normalized = F, credentials.mail = "marcelo.hurtado@inserm.fr", credentials.token = "734212f6ad77fc4eea2bdb502792f294")
deconv_Van = compute.deconvolution(counts.norm.Van, normalized = F, credentials.mail = "marcelo.hurtado@inserm.fr", credentials.token = "734212f6ad77fc4eea2bdb502792f294")
deconv = rbind(deconv_LP, deconv_Van)
```

PCA analysis with deconvolution
```{r}
compute_pca_analysis(t(deconv[rownames(traitData_early),]), traitData_early, 'Batch')
```
