---
title: "LP_early_analysis"
author: "Marcelo Hurtado"
output: html_document
date: "2024-08-03"
---

Set up environment
```{r setup, include=FALSE}
set.seed(123) #For reproducibility
source("src/environment_set.R") #Load functions and packages
```

Load data
```{r}
counts = read.csv("Counts_LungPredict.csv", row.names = 1) 
traitData = read.csv("ClinicalData_LungPredict.csv", row.names = 1)
```

Keep only ENSEMBL IDs
```{r}
m2 <- do.call(rbind, strsplit(rownames(counts), split="_", fixed = TRUE))
counts <- as.matrix(counts)
rownames(counts) = m2[,1]
```

Extract only coding genes
```{r}
library(biomaRt)

mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

gene_ids <- rownames(counts)

# Get gene type annotations
annotations <- getBM(attributes = c("ensembl_gene_id", "gene_biotype"), 
                     filters = "ensembl_gene_id", 
                     values = gene_ids, 
                     mart = mart)

# Filter annotations to keep only coding genes
coding_genes <- annotations %>%
  filter(gene_biotype == "protein_coding") %>%
  pull(ensembl_gene_id)
 
# Subset the count matrix to keep only coding genes
counts <- counts[rownames(counts) %in% coding_genes, ]
```

Change to SYMBOL
```{r}
# Extract gene symbols 
entrz <- AnnotationDbi::select(org.Hs.eg.db, keys = rownames(counts), columns = "SYMBOL", keytype = "ENSEMBL") 

# Choose first gene symbol whenever one ENSEMBL ID have multiple Gene SYMBOLS
entrz <- entrz %>%
  group_by(ENSEMBL) %>%
  slice(1) %>%
  ungroup()

# Final count matrix
counts = counts %>%
  data.frame() %>%
  mutate(Genes_symbol = entrz$SYMBOL) %>%
  na.omit(.) %>% #Remove genes with NO gene symbol
  remove_rownames() %>%
  distinct(Genes_symbol, .keep_all = TRUE) %>% #Keep only first gene symbol if duplicated
  column_to_rownames("Genes_symbol") 

```

EARLY STAGE
```{r}
traitData = traitData %>%
  filter(Stages_simplified %in% c("I", "II")) %>%
  dplyr::select(-Metastatic)

counts = counts[,rownames(traitData)]
```

Data normalization (if needed)
```{r}
counts.norm = ADImpute::NormalizeTPM(counts, log=T) 
```

Deconvolution

- Perform deconvolution using 6 methods and 10 signatures (GEMDeCan). It needs as input raw counts (genes as SYMBOL) and it will ask for your credentials to run CBSX
```{r}
deconv = compute.deconvolution(counts.norm, normalized = F, credentials.mail = "marcelo.hurtado@inserm.fr", credentials.token = "734212f6ad77fc4eea2bdb502792f294")
```

TFs inference
```{r}
tfs = compute.TFs.activity(counts.norm)
```

CellTFusion
```{r}
network = compute.WTCNA(tfs, corr_mod = 0.9, clustering.method = "ward.D2", minMod = 50) 
compute.metada.association(network[[1]], traitData, pval = 0.05, width = 10) 
pathways = compute.pathway.activity(counts.norm)
compute.modules.relationship(network[[1]], pathways, "Pathways_Progeny-TFs_Modules", vertical = T, width = 9, height = 5)
dt = compute.deconvolution.analysis(deconv, corr = 0.7, seed = 123) 
compute.modules.relationship(network[[1]], dt[[1]], "Deconvolution-TFs_Modules", vertical = T, height = 30, width = 10, pval = 0.05)
corr_modules = compute.modules.relationship(network[[1]], dt[[1]], return = T)
tfs.modules.clusters = compute.TF.network.classification(network, pathways)
cell_dendrograms = identify.cell.groups(corr_modules, tfs.modules.clusters, height = 20) 
res_all = cell.groups.analysis(dt[[1]], tfs.module.network = network, cell.dendrograms = cell_dendrograms, cut.height = 5, height = 20)
source("src/immunoscores.R")
gold.standards <- compute_gold_standards(RNA.tpm = data.frame(counts.norm))
```

Cell groups annotations
```{r}
module_colors = paste0("(", paste(unique(network[[2]]), collapse = "|"), ")")
  
dendrograms_colors <- sapply(colnames(res_all[[1]]), function(x) {
  match <- regexpr(module_colors, x)
  if (match != -1) {
    return(regmatches(x, match))
  } else {
    return(NA)
  }
})

features_annot = res_all[[1]] %>%
  t() %>%
  data.frame() %>%
  mutate(Color = paste0("TF_", dendrograms_colors))

features_annotation = function(row_annotations){
  row_annot <- rowAnnotation(Modules = row_annotations$Color, 
                             
                             col = list(Modules = c("TF_brown" = "brown", "TF_turquoise" = "turquoise", "TF_green" = "green", "TF_blue" = "blue",
                                                    "TF_black" = "black", "TF_yellow" = "yellow", "TF_red" = "red")),
                             
                             annotation_legend_param = list(labels_gp = gpar(fontsize = 10), legend_width = unit(12, "cm"), 
                                                            legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))  
  
  return(row_annot)
}



```

Heatmap annotations
```{r}
annotations = cbind(traitData, gold.standards)

heatmap_annotation = function(annotations){
  require(circlize)
  greyscale <- grey.colors(10, rev = T)
  col_fun = colorRamp2(c(0, 0.5, 1), c(brewer.pal(9, "YlOrBr")[1], brewer.pal(9, "YlOrBr")[4], brewer.pal(9, "YlOrBr")[9]))
  
  ann_colors <- HeatmapAnnotation(
    Stage = annotations$Stages_simplified, 
    Sex = annotations$Gender,
    CYT = annotations$CYT,
    Roh_IS = annotations$Roh_IS,
    Davoli_IS = annotations$Davoli_IS,
    IFNy = annotations$IFNy,
    Ayers_expIS = annotations$Ayers_expIS,
    Tcell_inflamed = annotations$Tcell_inflamed,
    
    col = list(
      Stage = c("I" = greyscale[1], "II" = greyscale[4], "III" = greyscale[7], "IV" = greyscale[10]),
      Sex = c("M" = "blue", "F" = "red"),
      CYT = col_fun,
      Roh_IS = col_fun,
      Davoli_IS = col_fun,
      IFNy = col_fun,
      Ayers_expIS = col_fun,
      Tcell_inflamed = col_fun
    ),
    
    annotation_legend_param = list(labels_gp = gpar(fontsize = 10), legend_width = unit(12, "cm"), 
                                   legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12))
    
  )
  
  return(ann_colors)
  
}
```

Heatmap cell groups
```{r}
hc = hclust(dist(res_all[[1]]), method = "ward.D2")
dend_column = as.dendrogram(hc)

ha = heatmap_annotation(annotations)

ht1 = Heatmap(t(scale(res_all[[1]])), 
        border = T, cluster_columns = dend_column, 
        top_annotation = ha, column_gap = unit(8, "mm"), #right_annotation = features_annotation(features_annot),
        name = "Cell groups", clustering_method_rows = "ward.D2", clustering_method_columns = "ward.D2",
        column_dend_height = unit(5, "cm"), row_dend_width = unit(5, "cm"), 
        column_dend_reorder = T, row_dend_reorder = F,
        show_row_names = T, 
        show_heatmap_legend = T, 
        row_names_gp = gpar(fontsize = 16), 
        column_names_gp = gpar(fontsize =12), 
        width = unit(25, "cm"), height = unit(10, "cm"),
        heatmap_legend_param = list(labels_gp = gpar(fontsize = 12), legend_width = unit(12, "cm"), 
                                    legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))

pdf("Results/Heatmap_cells_groups", width = 25, height = 15)
draw(ht1, show_heatmap_legend = T, heatmap_legend_side = "left", annotation_legend_side = 'left')
dev.off()
```
Patient classification 
```{r}
sub_grp <- cutree(hc, k = 2)

traitData = traitData %>%
  mutate(Cluster = sub_grp)

library(factoextra)
pdf("Results/Dendrogram_samples", width = 14, height = 8)
par(mar = c(5, 1, 1, 1)) #bottom, left, top, right
plot(dend_column, cex = 0.6)
dendextend::rect.dendrogram(dend_column, k=2, border = 2:5)
dev.off()
```

Feature selection
```{r}
data = res_all[[1]] %>%
  mutate(target = traitData$Cluster)

cell_groups_selected = feature.selection.boruta(data, iterations = 100, file_name = "Early_stage", thres = 0.9)
```

Heatmap of selected cell groups 
```{r}
boruta_groups = res_all[[1]][,colnames(res_all[[1]])%in%cell_groups_selected$Confirmed]

hc2 = hclust(dist(t(boruta_groups)), method = "ward.D2")
dend_row = as.dendrogram(hc2)

ht1 = Heatmap(t(scale(boruta_groups)), 
        border = T, cluster_rows = dend_row, cluster_columns = dend_column, column_split = 2,
        top_annotation = heatmap_annotation(annotations), column_gap = unit(8, "mm"), 
        name = "Cell groups", clustering_method_rows = "ward.D2", clustering_method_columns = "ward.D2",
        column_dend_height = unit(5, "cm"), row_dend_width = unit(5, "cm"), 
        column_dend_reorder = T, row_dend_reorder = F,
        show_row_names = T, 
        show_heatmap_legend = T, 
        row_names_gp = gpar(fontsize = 16), 
        column_names_gp = gpar(fontsize =12), 
        width = unit(25, "cm"), height = unit(10, "cm"),
        heatmap_legend_param = list(labels_gp = gpar(fontsize = 12), legend_width = unit(12, "cm"), 
                                    legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))

pdf("Results/Heatmap_cells_groups_boruta", width = 25, height = 10)
draw(ht1, show_heatmap_legend = T, heatmap_legend_side = "left", annotation_legend_side = 'left')
dev.off()
```

Extract cell types per cluster
```{r}
par(mar = c(15, 2, 1, 4)) #bottom, left, top, right
plot(dend_row, cex = 0.6)
rect.hclust(hc2, k = 2, border = 2:5)

cell_subgr <- cutree(hc2, k = 2)

for (clust in 1:length(unique(cell_subgr))) {
  cells_clust = unique(unlist(unname(res_all[[2]][which(cell_subgr == clust)])))
  cells_names_cluster = extract_cells(cells_clust)
  print(cells_names_cluster)
  print("====================================================================")
}
```

PCA analysis
```{r}
#PCA 
pdf("Results/PCA_cell_groups.pdf", width = 8, height = 5)
factoextra::fviz_cluster(list(data = boruta_groups, cluster = traitData$Cluster))
dev.off()

#Biplot
res.pca <- prcomp(boruta_groups,  scale = F)
factoextra::fviz_pca_biplot(res.pca, label="var", habillage=traitData$Cluster, select.var = list(contrib = 4),
               addEllipses=TRUE, ellipse.level=0.75)

# Extract the loadings
loadings <- res.pca$rotation
contribution <- (loadings^2)*100
features = contribution %>%
  data.frame() %>%
  rownames_to_column("Features") %>%
  arrange(desc(PC1))

pdf("Results/Cell_groups_contribution_pca.pdf", width = 12, height = 8)
p = ggplot(features, aes(x = reorder(Features, -PC1), y = PC1)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Contribution of cell groups",
       x = "Feature",
       y = "Contribution (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=12))
print(p)
dev.off()
```
Check cell group composition
```{r}
top = 5
for (i in 1:top) {
  print(res_all[[2]][features$Features[i]])
  print("========================================================================")
}
```
TFs modules analysis

Heatmaps of TFs modules

TF module red
```{r}
module_name = "red"
ht1 = Heatmap(t(tfs[,network[[3]][[module_name]]]), 
        border = T, cluster_columns = dend_column, 
        top_annotation = ha, column_gap = unit(8, "mm"), 
        name = paste("TF module", module_name), clustering_method_rows = "ward.D2", clustering_method_columns = "ward.D2",
        column_dend_height = unit(5, "cm"), row_dend_width = unit(5, "cm"), 
        column_dend_reorder = T, row_dend_reorder = F,
        show_row_names = T, 
        show_heatmap_legend = T, 
        row_names_gp = gpar(fontsize = 12), 
        column_names_gp = gpar(fontsize =12), 
        width = unit(20, "cm"), height = unit(30, "cm"),
        heatmap_legend_param = list(labels_gp = gpar(fontsize = 12), legend_width = unit(12, "cm"), 
                                    legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))

pdf(paste0("Results/Heatmap_TF_module_", module_name), width = 15, height = 18)
draw(ht1, show_heatmap_legend = T, heatmap_legend_side = "left", annotation_legend_side = 'left')
dev.off()
```

TF module yellow
```{r}
module_name = "yellow"
ht1 = Heatmap(t(tfs[,network[[3]][[module_name]]]), 
        border = T, cluster_columns = dend_column, 
        top_annotation = ha, column_gap = unit(8, "mm"), 
        name = paste("TF module", module_name), clustering_method_rows = "ward.D2", clustering_method_columns = "ward.D2",
        column_dend_height = unit(5, "cm"), row_dend_width = unit(5, "cm"), 
        column_dend_reorder = T, row_dend_reorder = F,
        show_row_names = T, 
        show_heatmap_legend = T, 
        row_names_gp = gpar(fontsize = 12), 
        column_names_gp = gpar(fontsize =12), 
        width = unit(20, "cm"), height = unit(45, "cm"),
        heatmap_legend_param = list(labels_gp = gpar(fontsize = 12), legend_width = unit(12, "cm"), 
                                    legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))

pdf(paste0("Results/Heatmap_TF_module_", module_name), width = 15, height = 25)
draw(ht1, show_heatmap_legend = T, heatmap_legend_side = "left", annotation_legend_side = 'left')
dev.off()
```

TF module blue
```{r}
module_name = "blue"
ht1 = Heatmap(t(tfs[,network[[3]][[module_name]]]), 
        border = T, cluster_columns = dend_column, 
        top_annotation = ha, column_gap = unit(8, "mm"), 
        name = paste("TF module", module_name), clustering_method_rows = "ward.D2", clustering_method_columns = "ward.D2",
        column_dend_height = unit(5, "cm"), row_dend_width = unit(5, "cm"), 
        column_dend_reorder = T, row_dend_reorder = F,
        show_row_names = T, 
        show_heatmap_legend = T, 
        row_names_gp = gpar(fontsize = 12), 
        column_names_gp = gpar(fontsize =12), 
        width = unit(20, "cm"), height = unit(55, "cm"),
        heatmap_legend_param = list(labels_gp = gpar(fontsize = 12), legend_width = unit(12, "cm"), 
                                    legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))

pdf(paste0("Results/Heatmap_TF_module_", module_name), width = 15, height = 28)
draw(ht1, show_heatmap_legend = T, heatmap_legend_side = "left", annotation_legend_side = 'left')
dev.off()
```

TF module green
```{r}
module_name = "green"
ht1 = Heatmap(t(tfs[,network[[3]][[module_name]]]), 
        border = T, cluster_columns = dend_column, 
        top_annotation = ha, column_gap = unit(8, "mm"), 
        name = paste("TF module", module_name), clustering_method_rows = "ward.D2", clustering_method_columns = "ward.D2",
        column_dend_height = unit(5, "cm"), row_dend_width = unit(5, "cm"), 
        column_dend_reorder = T, row_dend_reorder = F,
        show_row_names = T, 
        show_heatmap_legend = T, 
        row_names_gp = gpar(fontsize = 12), 
        column_names_gp = gpar(fontsize =12), 
        width = unit(20, "cm"), height = unit(40, "cm"),
        heatmap_legend_param = list(labels_gp = gpar(fontsize = 12), legend_width = unit(12, "cm"), 
                                    legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))

pdf(paste0("Results/Heatmap_TF_module_", module_name), width = 15, height = 22)
draw(ht1, show_heatmap_legend = T, heatmap_legend_side = "left", annotation_legend_side = 'left')
dev.off()
```

TF module turquoise
```{r}
module_name = "turquoise"
ht1 = Heatmap(t(tfs[,network[[3]][[module_name]]]), 
        border = T, cluster_columns = dend_column, 
        top_annotation = ha, column_gap = unit(8, "mm"), 
        name = paste("TF module", module_name), clustering_method_rows = "ward.D2", clustering_method_columns = "ward.D2",
        column_dend_height = unit(5, "cm"), row_dend_width = unit(5, "cm"), 
        column_dend_reorder = T, row_dend_reorder = F,
        show_row_names = T, 
        show_heatmap_legend = T, 
        row_names_gp = gpar(fontsize = 12), 
        column_names_gp = gpar(fontsize =12), 
        width = unit(20, "cm"), height = unit(60, "cm"),
        heatmap_legend_param = list(labels_gp = gpar(fontsize = 12), legend_width = unit(12, "cm"), 
                                    legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))

pdf(paste0("Results/Heatmap_TF_module_", module_name), width = 15, height = 30)
draw(ht1, show_heatmap_legend = T, heatmap_legend_side = "left", annotation_legend_side = 'left')
dev.off()
```

TF module black
```{r}
module_name = "black"
ht1 = Heatmap(t(tfs[,network[[3]][[module_name]]]), 
        border = T, cluster_columns = dend_column, 
        top_annotation = ha, column_gap = unit(8, "mm"), 
        name = paste("TF module", module_name), clustering_method_rows = "ward.D2", clustering_method_columns = "ward.D2",
        column_dend_height = unit(5, "cm"), row_dend_width = unit(5, "cm"), 
        column_dend_reorder = T, row_dend_reorder = F,
        show_row_names = T, 
        show_heatmap_legend = T, 
        row_names_gp = gpar(fontsize = 12), 
        column_names_gp = gpar(fontsize =12), 
        width = unit(20, "cm"), height = unit(25, "cm"),
        heatmap_legend_param = list(labels_gp = gpar(fontsize = 12), legend_width = unit(12, "cm"), 
                                    legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))

pdf(paste0("Results/Heatmap_TF_module_", module_name), width = 15, height = 15)
draw(ht1, show_heatmap_legend = T, heatmap_legend_side = "left", annotation_legend_side = 'left')
dev.off()
```

TF module brown
```{r}
module_name = "brown"
ht1 = Heatmap(t(tfs[,network[[3]][[module_name]]]), 
        border = T, cluster_columns = dend_column, 
        top_annotation = ha, column_gap = unit(8, "mm"), 
        name = paste("TF module", module_name), clustering_method_rows = "ward.D2", clustering_method_columns = "ward.D2",
        column_dend_height = unit(5, "cm"), row_dend_width = unit(5, "cm"), 
        column_dend_reorder = T, row_dend_reorder = F,
        show_row_names = T, 
        show_heatmap_legend = T, 
        row_names_gp = gpar(fontsize = 12), 
        column_names_gp = gpar(fontsize =12), 
        width = unit(20, "cm"), height = unit(50, "cm"),
        heatmap_legend_param = list(labels_gp = gpar(fontsize = 12), legend_width = unit(12, "cm"), 
                                    legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))

pdf(paste0("Results/Heatmap_TF_module_", module_name), width = 15, height = 25)
draw(ht1, show_heatmap_legend = T, heatmap_legend_side = "left", annotation_legend_side = 'left')
dev.off()
```

Number of TFs per module
```{r}
# Calculate the number of genes in each module
module_sizes <- sapply(network[[3]], length)

# Convert to a data frame for easier plotting
module_df <- data.frame(
  module = names(module_sizes),
  size = module_sizes
)

# Improved bar plot with text labels and module names adjusted
p = ggplot(module_df, aes(x = size, y = module, fill = module)) +
  geom_bar(stat = "identity", width = 0.6, color = "black") +  # Black borders for bars
  geom_text(aes(label = size), hjust = -0.2, size = 5) +  # Add text labels with a slight horizontal adjustment
  scale_fill_manual(values = names(module_sizes)) +  # Use module names as colors
  labs(
    x = "Frequency",
    y = "TF Module",
    title = "Number of TFs per module"
  ) +
  theme_minimal(base_size = 14) +  # Use a larger base size for readability
  theme(
    axis.title.x = element_text(margin = margin(t = 10), size = 14),  # No bold
    axis.title.y = element_text(margin = margin(r = 10), size = 14),  # No bold
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 12, color = "black", hjust = 1),
    plot.title = element_text(hjust = 0, size = 16),  # No bold
    legend.position = "none",  # Remove legend, as the bars themselves are self-explanatory
    panel.grid.major.y = element_blank(),  # Clean up horizontal grid lines
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "grey80", size = 0.5)  # Subtle grid for x-axis
  ) +
  coord_cartesian(clip = "off")  # Ensure labels are not cut off

pdf("Results/TF_module_frequency", width = 10, height = 10)
print(p)
dev.off()
```

Hub genes
```{r}
hub_tfs = identify_hub_TFs(t(tfs), network[[1]], network[[2]], MM_thresh = 0.8, degree_thresh = 0.9)
```

Heatmap hub genes

TFs modules annotations
```{r}
data = hub_tfs[[2]]
data$Module = paste0("TF_", data$Module)
row_annot <- rowAnnotation(Modules = data$Module, 
                           col = list(Modules = c("TF_brown" = "brown", "TF_turquoise" = "turquoise", "TF_green" = "green", "TF_blue" = "blue",
                                                    "TF_black" = "black", "TF_yellow" = "yellow", "TF_red" = "red")),
                           annotation_legend_param = list(labels_gp = gpar(fontsize = 10), legend_width = unit(12, "cm"), 
                                                            legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))  
```

Heatmap
```{r}
hub = unname(unlist(hub_tfs[["hubGenes"]]))
ht1 = Heatmap(t(tfs[,hub]), 
        border = T, cluster_columns = dend_column, 
        top_annotation = ha, column_gap = unit(8, "mm"), right_annotation = row_annot,
        name = "Hub TFs per module", clustering_method_rows = "ward.D2", clustering_method_columns = "ward.D2",
        column_dend_height = unit(5, "cm"), row_dend_width = unit(5, "cm"), 
        column_dend_reorder = T, row_dend_reorder = F,
        show_row_names = T, 
        show_heatmap_legend = T, 
        row_names_gp = gpar(fontsize = 12), 
        column_names_gp = gpar(fontsize =12), 
        width = unit(20, "cm"), height = unit(10, "cm"),
        heatmap_legend_param = list(labels_gp = gpar(fontsize = 12), legend_width = unit(12, "cm"), 
                                    legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))

pdf("Results/Heatmap_hubgenes", width = 15, height = 10)
draw(ht1, show_heatmap_legend = T, heatmap_legend_side = "left", annotation_legend_side = 'left')
dev.off()
```

Enrichment of TFs modules
```{r}
compute.modules.enrichment(counts.norm, hub_tfs)
```







====================================================================

Classify patients between top features
```{r}
traitData = traitData %>%
  mutate(Cluster1 = ifelse(res_all[[1]][,features$Features[1]] >= summary(res_all[[1]][,features$Features[1]])[5], 'High_1', 
                           ifelse(res_all[[1]][,features$Features[1]] < summary(res_all[[1]][,features$Features[1]])[2], "Low_1", "na")),
         Cluster2 = ifelse(res_all[[1]][,features$Features[2]] >= summary(res_all[[1]][,features$Features[1]])[5], 'High_2', 
                           ifelse(res_all[[1]][,features$Features[2]] < summary(res_all[[1]][,features$Features[1]])[2], "Low_2", "na")))
```

Differential expression analysis
```{r}
traitData$Cluster1 = factor(traitData$Cluster1)
traitData$Cluster2 = factor(traitData$Cluster2)

#Cluster 1
deseq2Data <- DESeqDataSetFromMatrix(countData=round(counts), colData=traitData, design= ~ Cluster1)
deseq2Data <- deseq2Data[rowSums(counts(deseq2Data)) > 5, ]
deseq2Data <- DESeq(deseq2Data)
resultsNames(deseq2Data)
deseq2Results <- results(deseq2Data, contrast = c("Cluster1", "Low_1", "High_1")) 
deseq2Results = data.frame(deseq2Results)
sigGenes <- rownames(deseq2Results[which(deseq2Results$padj <= .05 & abs(deseq2Results$log2FoldChange) > 2),])

library(EnhancedVolcano)
pdf("Results/DEG_Volcano_LP_cluster1", width = 10, height = 8)
EnhancedVolcano(deseq2Results, lab = rownames(deseq2Results), x = 'log2FoldChange', y = 'padj')
dev.off()

#Enrichment analysis
entrz <- AnnotationDbi::select(org.Hs.eg.db, keys = sigGenes, columns = "ENTREZID", keytype = "SYMBOL") #Change to EntrezID

kk <- enrichKEGG(gene         = entrz$ENTREZID,
                 organism     = 'hsa',
                 pvalueCutoff = 0.05)

pdf("Results/Enrichment_result_LP_cluster_1", width = 6, height = 7)
dotplot(kk, showCategory=35) + ggtitle("ORA - KEGG pathways")
dev.off()

#Cluster 2
deseq2Data <- DESeqDataSetFromMatrix(countData=round(counts), colData=traitData, design= ~ Cluster2)
deseq2Data <- deseq2Data[rowSums(counts(deseq2Data)) > 5, ]
deseq2Data <- DESeq(deseq2Data)
resultsNames(deseq2Data)
deseq2Results <- results(deseq2Data, contrast = c("Cluster2", "Low_2", "High_2")) 
deseq2Results = data.frame(deseq2Results)
sigGenes <- rownames(deseq2Results[which(deseq2Results$padj <= .05 & abs(deseq2Results$log2FoldChange) > 2),])

library(EnhancedVolcano)
pdf("Results/DEG_Volcano_LP_cluster2", width = 10, height = 8)
EnhancedVolcano(deseq2Results, lab = rownames(deseq2Results), x = 'log2FoldChange', y = 'padj')
dev.off()

#Enrichment analysis
entrz <- AnnotationDbi::select(org.Hs.eg.db, keys = sigGenes, columns = "ENTREZID", keytype = "SYMBOL") #Change to EntrezID

kk <- enrichKEGG(gene         = entrz$ENTREZID,
                 organism     = 'hsa',
                 pvalueCutoff = 0.05)

pdf("Results/Enrichment_result_LP_cluster_2", width = 6, height = 7)
dotplot(kk, showCategory=35) + ggtitle("ORA - KEGG pathways")
dev.off()
```
