---
title: "Vanderbilt_analysis"
author: "Marcelo Hurtado"
date: "2024-08-03"
output: html_document
---

Load results from LP
```{r setup, include=FALSE}
set.seed(123)
source("src/environment_set.R")
```

Load data
```{r}
counts = read.csv("Counts_Vanderbilt.csv", row.names = 1) 
traitData = read.csv("ClinicalData_Vanderbilt.csv", row.names = 1)
```

```{r}
library(stringr)
rownames(counts) <- str_replace(rownames(counts),
                        pattern = ".[0-9]+$",
                        replacement = "")
```

Extract only coding genes
```{r}
library(biomaRt)

mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

gene_ids <- rownames(counts)

# Get gene type annotations
annotations <- getBM(attributes = c("ensembl_gene_id", "gene_biotype"), 
                     filters = "ensembl_gene_id", 
                     values = gene_ids, 
                     mart = mart)

# Filter annotations to keep only coding genes
coding_genes <- annotations %>%
  filter(gene_biotype == "protein_coding") %>%
  pull(ensembl_gene_id)

counts = counts[rownames(counts) %in% coding_genes,]

# Change to gene SYMBOL
entrz <- AnnotationDbi::select(org.Hs.eg.db, keys = rownames(counts), columns = "SYMBOL", keytype = "ENSEMBL") %>%
  distinct(SYMBOL, .keep_all=T) %>%
  filter(ENSEMBL %in% rownames(counts)) %>%
  group_by(ENSEMBL) %>%
  slice(1) %>%
  ungroup()

counts = counts %>%
  rownames_to_column("Genes") %>%
  filter(Genes %in% entrz$ENSEMBL) %>%
  mutate(Symbol = entrz$SYMBOL) %>%
  na.omit() %>%
  remove_rownames() %>%
  column_to_rownames("Symbol") %>%
  dplyr::select(-Genes)

# Get gene type annotations
annotations_symbol <- getBM(attributes = c("hgnc_symbol", "gene_biotype"),
                     filters = "hgnc_symbol",
                     values = rownames(counts),
                     mart = mart)

# Filter annotations to keep only coding genes
coding_genes <- annotations_symbol %>%
  filter(gene_biotype == "protein_coding") %>%
  pull(hgnc_symbol)

counts = counts[rownames(counts) %in% coding_genes,]
```

Early Stage only (I, II) ~ 70 samples
```{r}
traitData = traitData %>%
  filter(Stages_simplified %in% c("Stage_I", "Stage_II")) %>%
  dplyr::select(-pt_ID, -Prior_Cancer_Type) %>%
  mutate(Stages_simplified = case_when(
    Stages_simplified == "Stage_I" ~ "I",
    Stages_simplified == "Stage_II" ~ "II",
    TRUE ~ Stages_simplified  # Keeps the original value if none of the conditions are met
  ))

counts = counts[,rownames(traitData)]
```

Rename Vanderbilt features to not replace them with the LP results
```{r}
counts_Van = counts
traitData_Van = traitData
# Upload LP early results
load("~/Documents/CellTFusion/user_projects/LP_early_results.RData")
```

Data normalization (if needed)
```{r}
counts.norm.Van = ADImpute::NormalizeTPM(counts_Van, log=T) 
```

PCA analysis
```{r}
traitData_Van$Batch = as.factor(traitData_Van$Batch)
compute_pca_analysis(counts.norm.Van, traitData_Van, 'Batch')
```

Perform analysis independently for each batch
```{r}
samples_1 = traitData_Van %>%
  filter(Batch == "1") %>%
  rownames_to_column("Samples") %>%
  pull(Samples)

samples_2 = traitData_Van %>%
  filter(Batch == "2") %>%
  rownames_to_column("Samples") %>%
  pull(Samples)

counts.norm.Van.1 = counts.norm.Van[,colnames(counts.norm.Van)%in%samples_1]
counts.norm.Van.2 = counts.norm.Van[,colnames(counts.norm.Van)%in%samples_2]
```

Deconvolution

- Perform deconvolution using 6 methods and 10 signatures (GEMDeCan). It needs as input raw counts (genes as SYMBOL) and it will ask for your credentials to run CBSX
```{r}
deconv_Van_1 = compute.deconvolution(counts.norm.Van.1, normalized = F, credentials.mail = "marcelo.hurtado@inserm.fr", credentials.token = "734212f6ad77fc4eea2bdb502792f294")
deconv_Van_2 = compute.deconvolution(counts.norm.Van.2, normalized = F, credentials.mail = "marcelo.hurtado@inserm.fr", credentials.token = "734212f6ad77fc4eea2bdb502792f294")
```

1. TFs inference
```{r}
tfs_Van_1 = compute.TFs.activity(counts.norm.Van.1)
tfs_Van_2 = compute.TFs.activity(counts.norm.Van.2)
```

Compute immunoscores
```{r}
source("src/immunoscores.R")
gold.standards_Van_1 <- compute_gold_standards(RNA.tpm = data.frame(counts.norm.Van.1))
gold.standards_Van_2 <- compute_gold_standards(RNA.tpm = data.frame(counts.norm.Van.2))
```

Join features from both batches
```{r}
deconv_Van = rbind(deconv_Van_1, deconv_Van_2)
tfs_Van = rbind(tfs_Van_1, tfs_Van_2)
gold.standards_Van = rbind(gold.standards_Van_1, gold.standards_Van_2)
```

PCA analysis
```{r}
compute_pca_analysis(t(deconv_Van), traitData_Van, 'Batch')
compute_pca_analysis(t(tfs_Van), traitData_Van, 'Batch')
compute_pca_analysis(t(gold.standards_Van), traitData_Van, 'Batch')
```

Batch cannot be removed from deconvolution (remove with ComBat_seq)
```{r}
counts_Van <- sva::ComBat_seq(as.matrix(counts_Van), batch=traitData_Van$Batch)
counts.norm.Van = ADImpute::NormalizeTPM(counts_Van, log=T)
deconv_Van = compute.deconvolution(counts.norm.Van, normalized = F, credentials.mail = "marcelo.hurtado@inserm.fr", credentials.token = "734212f6ad77fc4eea2bdb502792f294")
compute_pca_analysis(t(deconv_Van), traitData_Van, 'Batch')
```
Compute cell signatures from LP in Early stage samples Vanderbilt
```{r}
vanderbilt_groups = compute_cell_groups_signatures(dt, network, res_all, features = cell_groups_selected$Confirmed, deconv_Van, tfs_Van)
```

Heatmap annotation
```{r}
annotations = cbind(traitData_Van, gold.standards_Van)

heatmap_annotation = function(annotations){
  require(circlize)
  greyscale <- grey.colors(10, rev = T)
  col_fun = colorRamp2(c(0, 0.5, 1), c(brewer.pal(9, "YlOrBr")[1], brewer.pal(9, "YlOrBr")[4], brewer.pal(9, "YlOrBr")[9]))
  
  ann_colors <- HeatmapAnnotation(
    Stage = annotations$Stages_simplified, 
    Sex = annotations$Gender,
    CYT = annotations$CYT,
    Roh_IS = annotations$Roh_IS,
    Davoli_IS = annotations$Davoli_IS,
    IFNy = annotations$IFNy,
    Ayers_expIS = annotations$Ayers_expIS,
    Tcell_inflamed = annotations$Tcell_inflamed,
    
    col = list(
      Stage = c("I" = greyscale[1], "II" = greyscale[4], "III" = greyscale[7], "IV" = greyscale[10]),
      Sex = c("Male" = "blue", "Female" = "red"),
      CYT = col_fun,
      Roh_IS = col_fun,
      Davoli_IS = col_fun,
      IFNy = col_fun,
      Ayers_expIS = col_fun,
      Tcell_inflamed = col_fun
    ),
    
    annotation_legend_param = list(labels_gp = gpar(fontsize = 10), legend_width = unit(12, "cm"), 
                                   legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12))
    
  )
  
  return(ann_colors)
  
}

```

Heatmap of cell groups
```{r}
hc_van = hclust(dist(vanderbilt_groups), method = "complete")
dend_column = as.dendrogram(hc_van)

ha = heatmap_annotation(annotations)

ht1 = Heatmap(t(scale(vanderbilt_groups)), column_split = 2,
        border = T, cluster_columns = dend_column, cluster_rows = dend_row,
        top_annotation = ha, column_gap = unit(8, "mm"), 
        name = "Cell groups", clustering_method_rows = "ward.D2", clustering_method_columns = "complete",
        column_dend_height = unit(5, "cm"), row_dend_width = unit(5, "cm"), 
        column_dend_reorder = T, row_dend_reorder = F,
        show_row_names = T, 
        show_heatmap_legend = T, 
        row_names_gp = gpar(fontsize = 16), 
        column_names_gp = gpar(fontsize =12), 
        width = unit(25, "cm"), height = unit(10, "cm"),
        heatmap_legend_param = list(labels_gp = gpar(fontsize = 12), legend_width = unit(12, "cm"), 
                                    legend_heigh = unit(12, "cm"), title_gp = gpar(fontsize = 12)))

pdf("Results/Heatmap_cells_groups_boruta", width = 30, height = 15)
draw(ht1, show_heatmap_legend = T, heatmap_legend_side = "left", annotation_legend_side = 'left')
dev.off()
```

Patient classification
```{r}
traitData_Van = traitData_Van %>%
  mutate(Cluster = cutree(dend_column, k = 2))

library(factoextra)
pdf("Results/Dendrogram_samples", width = 14, height = 8)
par(mar = c(5, 1, 1, 1)) #bottom, left, top, right
plot(dend_column, cex = 0.6)
dendextend::rect.dendrogram(dend_column, k=2, border = 2:5)
dev.off()
```

PCA analysis
```{r}
#PCA 
pdf("Results/PCA_cell_groups.pdf", width = 8, height = 5)
factoextra::fviz_cluster(list(data = vanderbilt_groups, cluster = traitData_Van$Cluster))
dev.off()

#Biplot
res.pca <- prcomp(vanderbilt_groups,  scale = F)
factoextra::fviz_pca_biplot(res.pca, label="var", habillage=traitData_Van$Cluster, select.var = list(contrib = 4),
               addEllipses=TRUE, ellipse.level=0.75)

# Extract the loadings
loadings <- res.pca$rotation
contribution <- (loadings^2)*100
features = contribution %>%
  data.frame() %>%
  rownames_to_column("Features") %>%
  arrange(desc(PC1))

pdf("Results/Cell_groups_contribution_pca.pdf", width = 12, height = 8)
p = ggplot(features, aes(x = reorder(Features, -PC1), y = PC1)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Contribution of cell groups",
       x = "Feature",
       y = "Contribution (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size=12))
print(p)
dev.off()
```
Check cell group composition
```{r}
top = 5
for (i in 1:top) {
  print(res_all[[2]][features$Features[i]])
  print("========================================================================")
}
```

Differential expression analysis
```{r}
traitData_Van$Cluster = factor(traitData_Van$Cluster)
deseq2Data <- DESeqDataSetFromMatrix(countData=round(counts_Van), colData=traitData_Van, design= ~Cluster)
deseq2Data <- deseq2Data[rowSums(counts(deseq2Data)) > 5, ]
deseq2Data <- DESeq(deseq2Data)
resultsNames(deseq2Data)
deseq2Results <- results(deseq2Data, contrast = c("Cluster", "2", "1")) #Only between cluster 1 and 2
deseq2Results = data.frame(deseq2Results)
sigGenes <- rownames(deseq2Results[which(deseq2Results$padj <= .05 & abs(deseq2Results$log2FoldChange) > 1),])
foldchange <- deseq2Results[which(deseq2Results$padj <= .05 & abs(deseq2Results$log2FoldChange) > 1),]$log2FoldChange
names(foldchange) = sigGenes
library(EnhancedVolcano)
pdf("Results/DEG_Volcano_Vanderbilt", width = 10, height = 8)
EnhancedVolcano(deseq2Results, lab = rownames(deseq2Results), x = 'log2FoldChange', y = 'padj', labSize = 6)
dev.off()
```

Enrichment analysis
```{r}
entrz <- AnnotationDbi::select(org.Hs.eg.db, keys = sigGenes, columns = "ENTREZID", keytype = "SYMBOL") #Change to EntrezID

kk <- enrichKEGG(gene         = entrz$ENTREZID,
                 organism     = 'hsa',
                 pvalueCutoff = 0.01)

pdf("Results/Enrichment_result_Vanderbilt_dotplot", width = 6, height = 12)
dotplot(kk, showCategory=35) + ggtitle("ORA - KEGG pathways")
dev.off()
```

Other representations
```{r}
edox <- setReadable(kk, 'org.Hs.eg.db', 'ENTREZID')

pdf("Results/Enrichment_result_Vanderbilt_cnetplot", width = 10, height = 8)
cnetplot(edox, showCategory = c("Th1 and Th2 cell differentiation", "Natural killer cell mediated cytotoxicity",
                                "Antigen processing and presentation", "Cytokine−cytokine receptor interaction",
                                "Th17 cell differentiation"), color.params = list(foldChange = foldchange))
dev.off()

edox2 <- pairwise_termsim(edox)
pdf("Results/Enrichment_result_Vanderbilt_treeplot", width = 12, height = 10)
treeplot(edox2, showCategory = 20, hclust_method = "ward.D2")
dev.off()
```

```{r}
library(KEGGREST)
names(foldchange) = entrz$ENTREZID
data(bods, package = "pathview")
NK_cyto = pathview::pathview(gene.data = foldchange, pathway.id = 'hsa04650', species = "hsa", kegg.native = TRUE, out.suffix = "NK_cyto")
```
